---
title: "Flood detection using satellite data"
subtitle: "Introduction"
author: "Krzysztof Dyba"
format:
  html:
    toc: true
    toc-title: " "
    embed-resources: true
    code-links:
      - text: Repository
        icon: github
        href: https://github.com/kadyb/ogh2026
---

```{r}
#| message: false

library("terra") # 1.9.34
library("rstac") # 1.0.1
```

# Introduction

In September 2024, a [massive flood occurred](https://en.wikipedia.org/wiki/2024_Central_European_floods)
in Central Europe caused by record rainfall from the Genoa low. It is estimated
that the total losses could have exceeded 2 billion euros. For reference, take
a look at these photos from Wikimedia:
[Photo 1](https://upload.wikimedia.org/wikipedia/commons/0/0d/2024_Most_%C5%BCelazny_w_K%C5%82odzku_%287%29%2C_pow%C3%B3d%C5%BA.jpg),
[Photo 2](https://upload.wikimedia.org/wikipedia/commons/9/91/Prudnik%2C_2024.09.15_62.jpg),
[Photo 3](https://upload.wikimedia.org/wikipedia/commons/3/3a/2024_Pow%C3%B3d%C5%BA_w_K%C5%82odzku_%2816%29.jpg).

The situation was observed by satellites in space. The data they collect can be
used to monitor rainfall and floods in real time, which is very useful in crisis
management. The purpose of this analysis is to determine the extent of flooding
based on optical satellite imagery from [Sentinel-2](https://sentiwiki.copernicus.eu/web/sentinel-2).

# Analysis

## Data Acquisition

We will obtain satellite data from [SpatioTemporal Asset Catalogs](https://stacspec.org/)
(STAC) using the [rstac](https://github.com/brazil-data-cube/rstac) package.
This data will be sourced from the [Earth Search catalog](https://element84.com/earth-search/),
which is provided by Element 84.

We must use the `stac()` function and specify the URL to connect to the data
catalog.

```{r}
# define the STAC address
stac_source = stac("https://earth-search.aws.element84.com/v1")
stac_source
```

Next, we can search for the data we are interested in using the `stac_search()`
function and the specified criteria. The data collection (`collection`), spatial
extent (`bbox`), and temporal scope (`datetime`) will be important to us.
This is the fundamental part of the query. Furthermore, we can select specific
scenes based on specific attributes using the `ext_query()` function. In our
case, let's select one scene before the flood (September 8th) and one during
the flood (September 21st). Once we have the query ready, we can send it to the
server with the `post_request()` function.

::: {.callout-note appearance="simple"}
It will be very helpful to use the pipe operator (`|>`) to combine several
operations into one coherent sequence. We can use the keyboard shortcut
`CTRL` + `SHIFT` + `M`, but this requires first select the
`Use native pipe operator` option (`Code` > `Editing` tab) in RStudio.
:::

```{r}
# query with parameters
stac_source |>
  stac_search(
    collections = "sentinel-2-c1-l2a",
    bbox = c(17.4, 50.7, 17.7, 50.9), # xmin, ymin, xmax, ymax (WGS84)
    datetime = "2024-09-01T00:00:00Z/2024-09-30T00:00:00Z" # RFC 3339
  ) |>
  ext_query("datetime" %in% c("2024-09-08", "2024-09-21")) |>
  post_request() -> scenes
scenes
```

In response, we received an object that is a multi-level list of lists. We
can use the `View()` function to see its structure, which is quite complex.
The `lapply()` function will be useful for us, as it can iterate through
each element of the list and apply the same function.

We have found two scenes that meet our criteria. Let’s now select the spectral
bands that will be useful for our analysis:

| Band  | Name                        | Wavelength (nm) | Spatial Resolution |
|:-----:|:----------------------------|:---------------:|:------------------:|
| B2    | Blue                        | 458-523         | 10 m               |
| B3    | Green                       | 543-578         | 10 m               |
| B4    | Red                         | 650-680         | 10 m               |
| B8    | Near-Infrared (NIR)         | 785-899         | 10 m               |
| B11   | Shortwave Infrared (SWIR 1) | 1565-1655       | 20 m               |
| B12   | Shortwave Infrared (SWIR 2) | 2100-2280       | 20 m               |

: Selected spectral bands from Sentinel-2. {#tbl-sentinel}

We have two options for downloading data:

- Download the entire file to the disk using the `assets_download()` function
(from the `rstac` package).
- Download a subset of data representing a specific area of the image. This is
made possible by the [Cloud Optimized Geotiff](https://www.cogeo.org) (COG)
format and [GDAL Virtual File Systems](https://gdal.org/en/stable/user/virtual_file_systems.html).

Let's choose the second option, which requires providing a URL to access the
data. We can use the `assets_url()` function for this and we can indicate the
spectral bands that interest us using the `asset_names` argument. Importantly,
we want the returned URLs to be split by scene (rather than returned as a single
object), so we need to use `lapply()`.

```{r}
# get URLs for the images
bands = c("red", "green", "blue", "nir", "swir16", "swir22")
urls = lapply(scenes$features, assets_url, asset_names = bands)
```

We received two lists, each consisting of 6 URLs to images. Using `sapply()`
(returns a character vector directly instead of lists) and referring to the
`id` attribute, we can assign scene names to the `urls`.

```{r}
# assign scene names to the lists
names(urls) = sapply(scenes$features, "[[", "id")
```

The links are not sorted, which may cause errors related to the order of the
bands. So let's sort them using the `sort()` function.

```{r}
# sort links alphabetically
urls = lapply(urls, sort)
```

Finally, let's add the `/vsicurl/` prefix to each link to enable loading files
using HTTP range requests. We will use the `paste0()` function, but we can also
do this directly in `assets_url()` (set `append_gdalvsi` argument to `TRUE`).

```{r}
# add the "/vsicurl/" prefix
urls = lapply(urls, \(x) paste0("/vsicurl/", x))
```

## Data Processing

Let's define an area of interest that includes the flooded areas along the Nysa
Kłodzka River. We need to define this area as a spatial extent (also known as a
bounding box). The `ext()` function is used for this purpose. It requires four
coordinates (`xmin`, `xmax`, `ymin`, `ymax`) in a specific coordinate system.
We can check the coordinate system in the metadata or by using the `describe()`
function on the file. Our data is in the UTM Zone 33N coordinate system
(`EPSG:32633`).

```{r}
# define the spatial extent
# (xmin, xmax, ymin, ymax)
bbox = ext(c(669800, 689100, 5623100, 5641400))
```

The `rast()` function reads raster data. It requires the data source to be
specified as the first argument. In our case, these will be files located in
the cloud, and we will provide their addresses. In addition, as the `win`
argument, we will specify the `bbox` object containing the area of interest.
We will create two objects: `r_pre` containing images taken before the flood,
and `r_post` containing images taken during the flood.

```{r}
# read SWIR1 and SWIR2 bands (20 m)
r_pre = rast(urls[[2]][5:6], win = bbox)
r_post = rast(urls[[1]][5:6], win = bbox)
```

From the @tbl-sentinel table, we can see that the spectral bands have various
resolutions, which is problematic because we cannot overlay them (they have
different numbers of columns and rows). The solution is to bring them to the
same resolution. We will reduce the resolution from 10 m to 20 m (downsample)
using the `aggregate()` function. Let's load these four rasters (blue, green,
red, NIR bands) as temporary objects.

```{r}
# read R, G, B and NIR bands (10 m)
r1_tmp = rast(urls[[2]][1:4], win = bbox)
r2_tmp = rast(urls[[1]][1:4], win = bbox)
```

Let's downsample them by a factor of two, using the mean as the aggregation
function.

```{r}
#| output: false

# downsample rasters to 20 m
r1_tmp = aggregate(r1_tmp, fact = 2, fun = "mean")
r2_tmp = aggregate(r2_tmp, fact = 2, fun = "mean")
```

Given the same resolution, extent, and coordinate system, we can combine
(overlay) all the rasters. Simply use the standard `c()` function.

```{r}
# combine rasters
r_pre = c(r1_tmp, r_pre)
r_post = c(r2_tmp, r_post)
```

::: {.callout-note appearance="simple"}
Alternatively, instead of creating two separate raster objects, we could use
the `SpatRasterDataset` class, which represents multidimensional objects (with
time as the fourth dimension).
:::

For clarity, we can rename the layers (bands) from numbers to names by
overwriting the `names()` attribute.

```{r}
# rename layers
band_names = c("blue", "green", "red", "nir", "swir1", "swir2")
names(r_pre) = band_names
names(r_post) = band_names
```

The last step is to handle reflectance values that exceed the range of 0 to 1.
These are most often caused by atmospheric correction artifacts, sensor noise,
measurement errors or outliers. There are several ways to do this, but the most
commonly used method is the `clamp()` function, which allows us to remove values
that fall outside the defined range (`values = FALSE` argument) or to overwrite
them (`values = TRUE` argument). Deleting values may result in many empty spaces
on the map without data.

```{r}
# trim outliers
r_pre = clamp(r_pre, lower = 0, upper = 1, values = TRUE)
r_post = clamp(r_post, lower = 0, upper = 1, values = TRUE)
```

::: {.callout-note appearance="simple"}
Satellite data is typically stored as integers and must be scaled to a
floating-point range. For data from the Earth Search catalog, scaling factors
are stored in the raster metadata and are automatically applied upon loading.
See the `scoff()` function.
:::

## Data Visualization

We already have the data prepared, so let's see how it looks on the map. We can
use the simple `plot()` function to visualize a single band, but in our case,
a visual composition consisting of three images will be much more useful.
The composition can represent natural colors (red, green, blue), but can also
represent false colors, e.g. infrared bands (NIR, SWIR).

The `plotRGB()` function can be used to create a three-band composition.
We need to define the correct order of the raster layers (bands). It is also
usually necessary to enhance the contrast, for example, by applying linear
stretching (`stretch = "lin"` argument). Let's start by creating a composition
using natural colors (red, green, blue).

```{r}
#| layout-ncol: 2
#| fig-subcap:
#|   - "Before the flood"
#|   - "During the flood"

# red, green, blue bands
plotRGB(r_pre, r = 3, g = 2, b = 1, stretch = "lin")
plotRGB(r_post, r = 3, g = 2, b = 1, stretch = "lin")
```

Now let's create a false color composition consisting of the SWIR2, NIR and red
bands. Water almost completely absorbs near-infrared and short-wave infrared
radiation, making flooded areas clearly visible as dark contrasting surfaces
against a background of light-reflecting vegetation and soil.

```{r}
#| layout-ncol: 2
#| fig-subcap:
#|   - "Before the flood"
#|   - "During the flood"

# SWIR2, NIR, red bands
plotRGB(r_pre, r = 6, g = 4, b = 3, stretch = "lin")
plotRGB(r_post, r = 6, g = 4, b = 3, stretch = "lin")
```

## Spectral Index

Different objects absorb and reflect light in different ways depending on the
wavelength (a different value will be recorded in the green band and a different
one in the near infrared band). On this basis, we can calculate various spectral
indices; that is, by using appropriate ranges of the electromagnetic spectrum
and mathematical formulas, we can study the properties of objects. The most
common example of an index is the Normalized Difference Vegetation Index (NDVI),
which uses the red and near-infrared bands to assess the condition of vegetation.

Several indices have been developed for detecting water, with Normalized
Difference Water Index ([NDWI](https://doi.org/10.1080/01431169608948714))
and Modified Normalized Difference Water Index ([MNDWI](https://doi.org/10.1080/01431160600589179))
being the most commonly used.

Let's calculate the MNDWI, which uses the green and short-wave infrared (SWIR)
bands. The formula is as follows:

$$
\text{MNDWI} = \frac{\text{Green} - \text{SWIR}}{\text{Green} + \text{SWIR}}
$$

We must define the above formula as a function.

```{r}
# define MNDWI (Xu, 2006)
calculate_mndwi = function(green, swir) {
  mndwi = (green - swir) / (green + swir)
  names(mndwi) = "MNDWI"
  return(mndwi)
}
```

Then use this function for calculations by substituting the appropriate bands.

```{r}
# calculate MNDWI
mndwi_post = calculate_mndwi(r_post["green"], r_post["swir1"])
mndwi_pre = calculate_mndwi(r_pre["green"], r_pre["swir1"])
```

We can visualize the results on a map. The expected values for this index range
from -1 to 1, with high values indicating water extent.

```{r}
#| layout-ncol: 2
#| fig-subcap:
#|   - "MNDWI (before the flood)"
#|   - "MNDWI (during the flood)"

plot(mndwi_pre, range = c(-1, 1))
plot(mndwi_post, range = c(-1, 1))
```

## Change Detection

To determine the extent of the flood, we need to calculate the difference
between the current and initial state ($\text{MNDWI}_{\text{post}} - \text{MNDWI}_{\text{pre}}$).
Next, choose a specific threshold value that will indicate the presence of water.
The threshold value is selected empirically to minimize false positives
(e.g., shadows, wet soil or seasonal changes in vegetation), because spectral
characteristics can vary depending on weather conditions and sensor calibration.
There is no universal threshold that would work everywhere.

```{r}
# change in MNDWI between acquisitions
mndwi_diff = mndwi_post - mndwi_pre

# thresholding (TRUE if the condition is met)
water_mask = mndwi_diff > 0.4
```

```{r}
plot(water_mask, col = c("white", "blue"), main = "Water Mask")
```

After creating a mask with the flood extent, we can calculate its area using
the `expanse()` function. It is important to set the `byValue` argument to
`TRUE` to return exact raster values (in the below table, value 1 means `TRUE`,
0 means `FALSE`) and return the result in $km^2$.

```{r}
# calculate the area
expanse(water_mask, unit = "km", byValue = TRUE, transform = FALSE)
```

In the studied area, approximately 43 $km^2$ was affected by the flood.

## Export Results

We can save the results to disk using the `writeRaster()` function, specifying
the path and file format. Since we only have two values, we can set the datatype
to 8-bit (`datatype = "INT1U"`). By default, the data is saved using LZW
compression.

```{r}
#| eval: false

# save the results to disk
writeRaster(water_mask, "water_mask.tif", datatype = "INT1U")
```

# Exercise

1. Create an image composition consisting of the `swir2` (B12), `swir1` (B11)
and `nir` (B08) bands. Compare the images from before and during the flood.
2. Determine the area covered by the water based on the
[Automated Water Extraction Index](https://doi.org/10.1016/j.rse.2013.08.029)
(AWEI) and save the result to disk. How much area does the water cover?
To calculate AWEI, use the following formula:

$$
\text{AWEI} = 4 \times (\text{Green} - \text{SWIR}_1) - (0.25 \times \text{NIR} + 2.75 \times \text{SWIR}_2)
$$

