---
title: "Flood detection using satellite data"
subtitle: "Part 2: Machine learning"
author: "Krzysztof Dyba"
format:
  html:
    toc: true
    toc-title: " "
    embed-resources: true
    code-links:
      - text: Repository
        icon: github
        href: https://github.com/kadyb/ogh2026
---

```{r}
#| message: false

library("terra")      # 1.9.34
library("rpart")      # 4.1.27
library("rpart.plot") # 3.1.4
```

```{r}
# randomness seed
# for reproducibility of results
set.seed(1) 
```

# Introduction

In this part of the workshop, we will perform supervised classification to
detect the extent of flooding using an approach based on machine learning and
reference data. The trained model will be validated (to verify its accuracy)
and interpreted (to understand how it makes classification decisions).
Finally, we will make predictions for the entire area of interest.

# Analysis

## Data Loading

We'll start by loading the `points.gpkg` vector data in Geopackage format,
which is located in the `data` directory. The `vect()` function is used for
this.

```{r}
# load vector data
pts = vect("data/points.gpkg")
pts
```

We loaded a file consisting of 520 points and a single attribute representing
the land cover class (`class`). This attribute is character datatype, so we can
convert it to a factor representing categories using the `as.factor()` function.

```{r}
# convert variable to factor
pts$class = as.factor(pts$class)
```

We can retrieve unique classes by using the `levels()` function on this
attribute.

```{r}
# show unique classes
levels(pts$class)
```

We can also return the number of observations for an every category using
the `table()` function.

```{r}
# occurrence table
table(pts$class)
```

We can see that each class is represented by an equal number of observations.
In other words, this dataset is balanced.

Now let's load the raster data from the previous part of the workshop (the
rasters must be saved to disk, otherwise an error will appear at this point).

```{r}
# load raster data
r_pre = rast("data/r_pre.tif")
r_post = rast("data/r_post.tif")
```

Let's create visualization of reference points using an RGB composition as the
background. First, we need to call the `plotRGB()` function to create an RGB
composition, and then the `plot()` function to add points to it (set argument
`add` to `TRUE`). We also need to define an attribute from the vector layer
based on which colors will be assigned.

```{r}
# visualization
m = c(0, 0, 1.5, 0) # margins
colors = c("gold", "red", "darkgreen", "blue")

plotRGB(r_pre, r = 3, g = 2, b = 1, stretch = "lin",
        main = "Reference points", mar = m)
plot(pts, "class", col = colors, add = TRUE)
```

## Data Preparation

Typically, machine learning algorithms require input data in the form of a
matrix or a data frame. In supervised classification, a table consists of
explanatory (independent) variables and a predicted (dependent) variable.
In other words, the explanatory variables will represent the measured values,
and the predicted variable will contain reference data (labels/categories).

Based on the above, we need to get the band values for the reference points.
To do this, we will use the `extract()` function. We need to specify the raster
data as the first argument and the vector data as the second argument. We can
omit the point IDs from the table (`ID = FALSE` argument). By default, a data
frame will be returned as the result. Finally, we need to copy the `class`
attribute from the vector layer to the created data frame.

```{r}
# extract raster values based on points
df = extract(r_post, pts, ID = FALSE)

# copy attribute to data frame
df$class = pts$class
```

We can use the `View()` function to interactively view a data frame.

```{r}
#| eval: false

View(df)
```

The fact is that model outputs need to be verified. Various approaches are
used, such as holdout validation or cross-validation. In our case, we will
choose holdout validation, which requires dividing the input dataset into
two independent sets:

- a training set, on which the model will be trained,
- a test set, on which the model will be validated.

It is generally assumed that the training set constitutes 70% to 80% of the
observations, and the test set 20% to 30%, respectively.

From a technical perspective, the procedure is as follows. We calculate
the percentage of all observations that will constitute the training set,
e.g., 70%. Next, we randomly draw indices from the data frame using the
`sample()` function. Based on the drawn indices, we select the observations
and save them as the training set. We then create the test set by subtracting
those indices. 

```{r}
# divide into training and test sets
n = round(0.7 * nrow(df))
trainIndex = sample(nrow(df), size = n)
train = df[trainIndex, ]
test = df[-trainIndex, ]
```

## Model Training

For modeling, we will use a simple decision tree, which relies on classification
rules. The function is available in the `rpart` package under the same name.
The function requires defining a formula. We need to specify what is the modeled
variable and what are the explanatory variables. The tilde character indicates
dependency, e.g. `Y` is dependent on `X1` and `X2`. The plus sign is used to
combine subsequent variables. For instance:

```r
# Y depends on X1 and X2
Y ~ X1 + X2
```

If a dataset contains multiple explanatory variables, we don't have to list
them all. In that case, you can use a period to represent all variables except
the dependent variable. Then the simpler form: `Y ~ .`

In addition to the formula, we must also define a dataset (`data = train`) and
a method. In this case, the method will be classification (`method = "class"`).
We can also define the model's hyperparameters to improve classification
results. To simplify the model, let's set the model complexity (`cp` parameter)
from the default value of 0.01 to 0.05. In practice, this will result in a
shallower decision tree (fewer splits).

```{r}
# train decision tree
mdl_dt = rpart(class ~ ., data = train, method = "class",
               control = rpart.control(cp = 0.05))
```

## Model Validation

To evaluate the quality of a model, we need to use some kind of statistic
(metric). For classification, the simplest metric is accuracy, which is
calculated as follows:

$$
\text{Accuracy} = \frac{\text{TP} + \text{TN}}{\text{TP} + \text{TN} + \text{FP} + \text{FN}}
$$

Where:

* **TP** – True Positives
* **TN** – True Negatives
* **FP** – False Positives
* **FN** – False Negatives

According to the formula, we need to sum the correct classifications and divide
that sum by the total number of classifications.

```{r}
# define accuracy metric
accuracy = function(actual, predicted) {
  accuracy = sum(actual == predicted) / length(actual)
  return(accuracy)
}
```

Let's check how effective our trained model is using the test set. From the
test set, we exclude the column containing the reference class and make
predictions using the model. The `predict()` function is used for this.
The order of the arguments is important. If we are making a prediction for
a data frame, we specify the model as the first argument and the data frame
as the second argument. In addition, we must define the type of prediction,
which in this case will be `"class"`.

```{r}
# prediction for test set
pred = predict(mdl_dt, test[, -7], type = "class")
```

This gives us a vector containing the predicted classes. In the created
`accuracy()` function, we provide a vector with reference classes and a
vector with predicted classes.

```{r}
# calculate accuracy
accuracy(test$class, pred)
```

We achieved an accuracy of over 90%, which is a very good result. Next, we
can examine the classification results in detail using a confusion matrix.
This is a tabular summary showing which classes are misclassified by the model.
The upper triangle in the table shows the False Positive (type I error)
classifications, while the lower triangle shows the False Negative (type II error)
classifications. Ideally, most values should be on the diagonal of the table,
because they indicate correct classifications. This table can be created using
the `table()` function.

```{r}
# confusion matrix
table(Predicted = pred, Reference = test$class)
```

## Model Explanation

In most cases, machine learning models are black box models, and it is unclear
how they work. A decision tree, on the other hand, is a white box model, and
we can understand the classification rules.

```{r}
# plot decision tree
prp(mdl_dt)
```

From the figure above, we can see that two conditions are necessary for the
model to classify water:

1. Values in the SWIR2 band are less than 0.073.
2. Values in the NIR band are less than 0.14.

Let's plot these two conditions as dashed lines on a scatter plot of the
original data.

```{r}
colors = c("gold", "red", "darkgreen", "blue")

plot(df$nir, df$swir2, col = colors[df$class], pch = 19,
     main = "Water detection",
     xlab = "Near Infrared (NIR)", ylab = "Shortwave Infrared 2 (SWIR2)")
abline(h = 0.073, v = 0.136, lty = 2, lwd = 2) # threshold values
legend("topright", title = "Reference class", legend = levels(df$class),
       col = colors, pch = 19)
```

## Final Map

Let's make a prediction for the entire area. We will use the `predict()`
function again, but this time for raster data rather than a data frame.
In this case, we need to provide the raster data as the first argument
to the function and the model as the second. We will set the prediction
type again to classification. If there are missing values in the raster,
it is a good idea to set the `na.rm` argument to `TRUE`.

```{r}
# prediction for the entire area
pred_map = predict(r_post, mdl_dt, type = "class", na.rm = TRUE)
```

The prediction map has four classes (cropland, developed, forest, water). Let's
convert these classes into a binary raster representing a flood extent mask.
To do this, we need to use the `classify()` function to reclassify the values.
Reclassification is performed based on a reclassification matrix consisting of
two columns. The first column contains the old value, and the second column
contains the new value. Let's set the water class to 1 and the other classes
to 0.

```{r}
# reclassification matrix
rcl_mat = c(
  1, 0,  # cropland   -->  FALSE
  2, 0,  # developed  -->  FALSE
  3, 0,  # forest     -->  FALSE
  4, 1   # water      -->  TRUE
)
rcl_mat = matrix(rcl_mat, ncol = 2, byrow = TRUE)
```

```{r}
# reclassify raster values (binary mask)
water_mask = classify(pred_map, rcl_mat)
```

On the map, we can see noise consisting of many individual cells in homogeneous
areas (the "salt-and-pepper" effect). The reason for this is that classification
is performed for each cell independently, without taking its local neighborhood
into account. However, we can address this issue by smoothing the raster values
using a modal filter. We can do this by using the `focal()` function, defining
the size of the moving window (`w` argument), and specifying a function
(`fun = "modal"`). The larger the window size, the smoother the values will be.

```{r}
# smooth raster values
water_mask = focal(water_mask, w = 3, fun = "modal", expand = TRUE)
```

Let's compare the water extent predicted by the model with a false color
composition (SWIR2, NIR and red bands).

```{r}
#| layout-ncol: 2
#| fig-subcap:
#|   - "Prediction"
#|   - "False color composition"

colors = c("white", "blue")
plot(water_mask, col = colors, legend = FALSE, axes = FALSE, mar = 0)

plotRGB(r_post, r = 6, g = 4, b = 3, stretch = "lin")
```

Finally, we can save the result to disk using the `writeRaster()` function.

```{r}
#| eval: false

# save the results to disk
writeRaster(water_mask, "water_mask_ml.tif", datatype = "INT1U")
```

# Exercise

1. Train and test an support vector machine (SVM) model from the e1071 package
(see `?svm`). Present the classification result on a map.

