---
title: "Analysis of land surface temperature based on thermal satellite data"
author: "Krzysztof Dyba"
format:
  html:
    toc: true
    toc-title: " "
    embed-resources: true
    code-links:
      - text: Repository
        icon: github
        href: https://github.com/kadyb/ogh2026
---

```{r}
#| message: false

library("terra") # 1.9.34
```

```{r}
# randomness seed
# for reproducibility of results
set.seed(1) 
```

# Introduction

In addition to standard optical bands, the [Landsat](https://science.nasa.gov/mission/landsat/)
satellite also has two thermal infrared bands that measure thermal radiation
(10.60 - 11.19 µm and 11.50 - 12.51 µm). The original thermal data are acquired
at a resolution of 100 m and upsampled to 30 m. Landsat data can be searched
for and downloaded for free from the [Earth Explorer](https://earthexplorer.usgs.gov/)
website.

[ESA WorldCover](https://esa-worldcover.org) is a global land cover classification
based on Sentinel-1 and Sentinel-2 data at a resolution of 10 m. The classification
was performed for 2020 and 2021, for which 11 classes were defined. The data can
be downloaded for free from, among others, the [WorldCover viewer](https://viewer.esa-worldcover.org/worldcover).

The purpose of this analysis is to assess the surface temperature of various
land cover classes. The study area covers Istanbul. The Landsat scene was
acquired on July 23, 2025 at 8:45 am. According to [Weather Underground](https://www.wunderground.com/),
the temperature at that time was about 32°C.

::: {.callout-note appearance="simple"}
Air temperature is not directly comparable with satellite-derived land surface
temperature. A sunlit roof, road, or bare field can have a much higher surface
temperature than the surrounding air.
:::

# Analysis

## Data Preparation

The necessary data can be found in the [Zenodo repository](https://zenodo.org/records/21922843).
We can access this data in two ways:

1. We can download them directly to the disk using the `download.file()`
function. In this case, it is also a good idea to increase the default timeout.
The data needs to be extracted from the archive using the `unzip()` function
after downloading. Let's choose this method as preferred.

```{r}
if (!dir.exists("data/landsat")) {
  
  # increase connection timeout
  options(timeout = 600)
  url = "https://zenodo.org/api/records/21922843/files-archive"
  
  # download archive from URL
  download.file(url, "data/landsat.zip", mode = "wb")
  
  # unzip archive
  unzip("data/landsat.zip", exdir = "data/landsat")

}
```

2. Alternatively, use [GDAL Virtual File Systems](https://gdal.org/en/stable/user/virtual_file_systems.html)
to access them remotely. This requires adding the prefix `/vsicurl/` to both
tif files. For example:

```{r}
#| eval: false

url_landcover = "https://zenodo.org/records/21922843/files/ESA_WorldCover.tif?download=1"
url_landcover = paste0("/vsicurl/", url_landcover)
landcover = rast(url_landcover)
```

### Satellite Data

Let's load the four bands representing surface reflectance (B2–B5) using the
`rast()` function and defining the indices of these bands in the `lyrs`
argument.

```{r}
# read surface reflectance bands (B2-B5)
landsat = rast("data/landsat/Landsat.tif", lyrs = 1:4)
```

Optionally, we can assign new names to the default names of raster bands (layers)
using the `names()` function. 

```{r}
# assign names to the bands
# names(landsat) = c("blue", "green", "red", "nir")
```

Let's also load the last band representing surface temperature (B10).

```{r}
# read thermal band (B10)
thermal = rast("data/landsat/Landsat.tif", lyrs = 5)
```

The downloaded data comes from Landsat's Collection 2 Level 2. Level 2 means
that the data has been atmospherically corrected and represents the estimated
surface reflectance and surface temperature. Please note that only one thermal
band (B10) is available in Level 2.

The data is stored as integers, so we need to convert it to floating-point
numbers using the appropriate formula. We must multiply the recorded values
by a scale factor and add an offset:

- For surface reflectance, the scale factor is $0.0000275$ and the offset is
$-0.2$.
- For surface temperature (B10 band), the scale factor is $0.00341802$ and the
offset is $149$. The converted values will be in Kelvin (K).

Let's scale the values of B2–B5 bands (`landsat` object) using these coefficients.

```{r}
# scale reflectance values
landsat = landsat * 0.0000275 - 0.2
```

After scaling, we need to correct the values that fall outside the standard
reflectance range of 0 to 1. We can use the `clamp()` function for this and
define minimum and maximum values. With `values = TRUE`, values below or
above the specified limits are replaced by the corresponding limit.

```{r}
# overwrite values outside the range
landsat = clamp(landsat, lower = 0, upper = 1, values = TRUE)
```

We can create a visualization consisting of three bands -- red, green, and blue.
We should also use the `stretch = "lin"` argument to increase the image's
contrast.

```{r}
# RGB composition
plotRGB(landsat, r = 3, g = 2, b = 1, stretch = "lin")
```

Now, let's apply scaling to the thermal data (keep in mind that the coefficients
have different values).

```{r}
# scale temperature values to Kelvin
thermal = thermal * 0.00341802 + 149
```

The result is expressed in Kelvin. For convenience, let's convert it to Celsius.

```{r}
# scale temperature to Celsius
thermal = thermal - 273.15
```

Let's see what the spatial variation in surface temperature looks like on the
map. For this visualization, we will use the `Plasma` color palette, which can
be generated using the `hcl.colors()` function.

```{r}
plasma_colors = hcl.colors(n = 40, palette = "Plasma")
plot(thermal, col = plasma_colors, main = "Surface temperature [°C]")
```

### Land Cover

```{r}
# read landcover raster
landcover = rast("data/landsat/ESA_WorldCover.tif")
```

```{r}
# read csv file with labels
df = read.csv("data/landsat/labels.csv")
```

After loading the data frame, we can assign appropriate labels to the raster
values. This is done using the `levels()` function, which requires the data
frame containing the IDs and labels.

```{r}
# assign labels to IDs
levels(landcover) = df
```

We can see that some classes do not appear at all in our area of interest.
To remove such classes, use the `droplevels()` function.

```{r}
# remove unnecessary classes
landcover = droplevels(landcover)
```

Both datasets have the same spatial extents (bounding boxes). However, in
the case of the satellite data, we can notice that values are missing in
the southern part of the scene (they were not recorded in that area). We
can transfer the `NA` mask from the satellite image to a land cover raster
using the `mask()` function.

```{r}
# copy the NA values from the satellite data to the landcover
landcover = mask(landcover, thermal)
```

```{r}
mar = c(0, 1.5, 0, 7.2)
plot(landcover, main = "Land cover", mar = mar)
```

## Thresholding

Using a simple threshold, let's identify the areas with the highest surface
temperatures. Let's consider the 80th percentile to be the threshold value.
To calculate this quantile, we must use the `global()` function and set the
following arguments: `fun = quantile` (without quotes), `probs = 0.80`,
and `na.rm = TRUE`. The result will be a data frame, which we can convert
to a vector using the `unlist()` function.

```{r}
# calculate the 80th percentile
q80 = global(thermal, fun = quantile, probs = 0.80, na.rm = TRUE)

# convert the data frame to a vector
q80 = unlist(q80, use.names = FALSE)
q80
```

```{r}
# find the warmest areas based on the percentile
high_temp = thermal > q80
```

As a result, we obtained a binary mask consisting of two values: 0 (FALSE)
if the condition was not met, and 1 (TRUE) if the condition was met.
To overlay this mask on the map, we need to replace the values of 0 with
`NA` to make those areas transparent. Otherwise, they will be obscured.

```{r}
# replace 0 with NA
high_temp[high_temp == 0] = NA
```

```{r}
mar = c(0, 0, 2, 0)
plotRGB(landsat, r = 3, g = 2, b = 1, stretch = "lin", mar = mar,
        main = "Surface temperature over 50°C")
plot(high_temp, col = "brown", legend = FALSE, add = TRUE)
```

The highest surface temperatures are recorded primarily in built-up areas and
on croplands (mainly bare fields without vegetation).

## Data Sampling

Let's draw a class-balanced sample from the land cover raster. Our sampling
design assumes that each class should occur as frequently as the others. To
achieve this, we must use stratified sampling. The `spatSample()` function is
used for sampling and supports various sampling methods. For stratified sampling,
we must set the `method` argument to `stratified`. The `size` argument specifies
how many observations will be returned for each class. Therefore, setting its
value to 1,000 will result in a total of 8,000 points. In addition, we specify
that a point layer will be returned (`as.points = TRUE` argument).

```{r}
# stratified sampling
points = spatSample(landcover, size = 1000, method = "stratified",
                    as.points = TRUE, values = FALSE)
```

Now that we have the points, we can retrieve values from the rasters using the
`extract()` function. We use `bind = TRUE` to append the extracted values to
this vector layer.

```{r}
# extract raster values by points
points = extract(landcover, points, bind = TRUE)
points = extract(thermal, points, bind = TRUE)
```

## Data Aggregation

At this point, we have a vector layer with two attributes -- land cover class
and surface temperature. We can then calculate descriptive statistics for each
class using the `aggregate()` function. For instance, let's calculate the mean
value.

```{r}
# average temperature for each class
agg = aggregate(B10 ~ label, data = points, FUN = "mean")
```

We can sort the results in descending order using the `order()` function on
the `B10` column (for descending order, it is easiest to place a minus sign
before the column name). Next, we can round the values to one decimal place
using the `round()` function and the `digits = 1` argument. We can also
modify the column names using the `colnames()` function.

```{r}
# descending sort
agg = agg[order(-agg$B10), ]

# round values
agg$B10 = round(agg$B10, digits = 1)

# rename columns
colnames(agg) = c("Class", "Value")
```

```{r}
#| echo: false

knitr::kable(agg, caption = "Average surface temperature [°C]", align = "c",
             row.names = FALSE)
```

## Comparing Groups

Naturally, the question arises as to whether the surface temperatures for these
classes actually differ. We can verify this in two ways by using:

- Visual data exploration, e.g., a box plot, which is used to present position
statistics, or a histogram, which is used to analyze the shape of the distribution.
- Statistical tests, which help determine whether the observed patterns are the
result of a real effect or could be due to chance.

Every statistical test is based on a logical process called hypothesis testing,
which generally works as follows:

1. Formulate hypotheses:
    - Null hypothesis ($H_0$) -- assumes no effect or no difference exists.
    - Alternative hypothesis ($H_1$) -- assumes there is an effect or difference.
2. Calculate a test statistic based on the sample data, which produces a
p-value (probability value). The p-value (a number between 0 and 1) indicates
how surprising the data are if we did not expect to find anything in the data
that deviates from the null hypothesis (no effect). A small p-value means the
observed result would be very unlikely under the null hypothesis, which gives
reason to doubt the null hypothesis.
3. Compare the p-value with the significance level (alpha; $\alpha$).
We must choose a threshold value ($\alpha$) that will be used to reject the
null hypothesis ($H_0$). It is generally accepted that the threshold value
is 0.05 or 0.01. Then:
    - p-value $\leq$ $\alpha$ -- we reject the null hypothesis and conclude that
  the effect is statistically significant.
    - p-value > $\alpha$ -- we cannot reject the null hypothesis.

Statistical tests are divided into two types: parametric tests and nonparametric
tests. Parametric tests are based on strict assumptions about the shape and
distribution of the population from which the data are drawn, whereas
nonparametric tests are based on very few assumptions or none at all.

Assumptions of parametric tests:

- Data is normally distributed (bell-shaped curve with few/no extreme outliers).
- Variance is roughly equal across groups (spread of values is similar between
  groups).
- Sample sizes are reasonably large (n $\geq$ 30).

### Visual Data Exploration

The `boxplot()` function will generate box plots taking classes into account,
while the `hist()` function will generate histograms. To create the histograms,
we need to use a loop and select the appropriate land cover class in each
iteration. It is also recommended to set the same range of values on the
X-axis so that different cases can be compared more easily.

```{r}
# boxplots
par(mar = c(5, 9, 1, 1))
boxplot(B10 ~ label, data = points, horizontal = TRUE, las = 1,
        xlab = "Surface temperature [°C]", ylab = NULL)
```

```{r}
# histograms on a single panel
par(mfrow = c(3, 3))
for (i in levels(points$label)) {
  
  # select appropriate class
  idx = points$label == i
  title = paste0(i, " [°C]")
  hist(points$B10[idx], xlab = NULL, main = title, xlim = c(25, 60))
  
}
```

What conclusions can be drawn from the figures above?

1. Some land cover classes have many outliers.
2. The differing widths of the boxes show that the classes have different
variances.
3. The histograms are not symmetric (some have tails). Additionally, the
medians are not centered within the boxes, and the whiskers are unequal,
indicating skewed distributions. Therefore, we can conclude that the condition
of normality is not met.

We can also confirm our deductions using statistical tests. For example,
the Shapiro-Wilk test (`shapiro.test()`) can be used to check for a normal
distribution, and Levene's test (`car::leveneTest()`) can be used to check
for homogeneity of variances.

### Statistical Testing

We assume that the requirements for parametric tests are not met, so we must
use a nonparametric test to compare the groups. In our case, we can use the
Kruskal-Wallis test. Our hypotheses are:

- Null Hypothesis ($H_0$): No difference between groups.
- Alternative Hypothesis ($H_1$): A real difference exists.

We set the significance level ($\alpha$) to 0.01. We will use the `kruskal.test()`
function to calculate this test.

```{r}
# Kruskal-Wallis test
kruskal.test(B10 ~ label, data = points)
```

The p-value is very low (below the selected threshold), so we can reject the
null hypothesis ($H_0$) and accept the alternative hypothesis ($H_1$) that
there are significant differences between the classes. However, we do not
know between which groups this difference is significant.

Therefore, in the next step, we must perform a post-hoc test to compare the
groups in pairs and identify the exact differences (again, using a nonparametric
version). For instance, we can use pairwise Wilcoxon tests implemented in the
`pairwise.wilcox.test()` function.

```{r}
# pairwise Wilcoxon tests
result = pairwise.wilcox.test(points$B10, points$label)
```

```{r}
#| echo: false

format_pvalue = function(x) {
  if (is.na(x)) return("-")
  if (x < 0.001) return("<0.001")
  return(as.character(round(x, 3)))
}

result$p.value = apply(result$p.value, MARGIN = c(1, 2), FUN = format_pvalue)
knitr::kable(result$p.value, caption = "Pairwise Wilcoxon Test", align = "c")
```

Similarly, if the p-value is less than the specified significance level, we
reject the null hypothesis and accept the alternative hypothesis, which states
that the difference is significant.

# Exercise

1. Is there a relationship between surface temperature and the normalized
difference vegetation index (NDVI)? Examine this relationship using a scatter
plot. To calculate NDVI, use the following formula:

$$
\text{NDVI} = \frac{\text{NIR} - \text{RED}}{\text{NIR} + \text{RED}}
$$

